Autopolyploid establishment through polygenic adaptation

### 2025-01-23 doc/paper/paper.tex
revisions

    \textbf{Keywords:} polyploidy, adaptation, establishment, inbreeding,
    infinitesimal model
\end{abstract}

@@

Hence, to explain the widespread occurrence of mixed-ploidy populations,
additional factors besides the continuous formation of polyploids through the
union of unreduced gametes need to be considered.
Firstly, chance establishment of tetraploids through drift could occur.
Indeed, the problem is somewhat analogous to the spread of underdominant
chromosomal rearrangements, where local establishment through genetic drift and
subsequent spreading in a subdivided population through local extinction and 
recolonization dynamics has been suggested as a plausible model
\citep{lande1985}.
However, MCE is quite strong in randomly mating mixed-ploidy populations, and
the population size has to be very small for local tetraploid establishment to
occur at an appreciable rate (\cite{rausch2005}, see also \cref{s-sec:mchain}).
Secondly, any form of \textit{prezygotic isolation} between cytotypes could
promote establishment of polyploid cytotypes by alleviating MCE.
Particularly relevant are assortative mating by cytotype  (for instance through
phenological differences across cytotypes, or differences in pollinators;
\cite{kolar2017}), self-fertilization \citep{rausch2005,novikova2023}, and
localized dispersal \citep{baack2005,kolar2017}.
Finally, selection may be invoked to explain the establishment of polyploids.
Tetraploids may have higher relative fitness than their diploid counterparts
due to reduced inbreeding depression \citep{ronfort1999}, or due to being
better adapted to (changing) environmental conditions \citep{vandepeer2021}.

@@

In generation $t$, $M(t)$ migrant individuals arrive on an island (the new
habitat) joining $N^\ast(t)$ resident individuals, where $M(t)$ is Poisson
distributed with mean $m$. 
The migrant individuals are assumed to be unrelated to the resident
individuals.
After migration in generation $t$, the $N(t) = N^\ast(t) + M(t)$ individuals
reproduce sexually, and the offspring thus produced survives until the next
generation with a probability determined by their trait value.
In the basic model, random selfing is allowed (but see below for a model with
self-incompatibility).
We assume the trait is under directional selection, with fitness $w(z) =
e^{\gamma(z - \theta)}$, where $\gamma$ is the intensity of directional
selection and $\theta$ is the trait value for which the growth rate of the
island population becomes positive.
Throughout the paper, we define establishment as reaching a population of size
100.

@@

We evaluate the accuracy of the autotetraploid infinitesimal model as an
approximation to the evolution of a quantitative trait determined by $L$
additive loci.
We find that the infinitesimal model with inbreeding generally yields accurate
predictions for the evolution of the genetic variance when the number of loci
is sufficiently large ($L \ge 100$, say, \cref{fig:vztet,s-fig:fininf}).
Furthermore, we confirm that, in the absence of double reduction, the decay in
genetic variance due to inbreeding after a time $t$ is well-predicted by
$e^{-t/4N}$ (\cref{fig:vztet}A), as expected from the results of \cite{arnold2012}.
As predicted, double reduction (i.e. $\alpha_4 > 0$) leads to an immediate
increase in genetic variance (as it increases the segregation variance), but
leads to accelerated inbreeding, causing faster loss of genetic variance in the
long term (\cref{fig:vztet,s-fig:fininf}).

Simulations for the mixed-ploidy model further confirm the correctness of our
infinitesimal approximation (\cref{s-fig:vz}), yielding accurate predictions
for the evolution of the phenotypic variance across ploidy levels and for
different assumptions regarding the scaling of allelic effects ($\beta$
parameters) and unreduced gamete formation ($\alpha$ parameters).

It is worth noting that, although inbreeding is slower in autotetraploids than
in diploids for the same population size, the tetraploid fraction of a
diploid-dominated mixed-ploidy population will have an equal or higher average
inbreeding coefficient (\cref{s-fig:f1}).
This is because in such a population, triploid and tetraploid individuals
mostly arise from gametes formed by diploid individuals, or by polyploid
individuals with very recent diploid ancestry (on average $1+u+2v$ generations ago
for tetraploids, and $1+\frac{2}{3}(u+2v)$ generations ago for triploids, see
\cref{s-sec:ttdip}).
As a result, the polyploid subpopulations will show an average relatedness
similar to that of the diploid population and not evolve as an isolated
higher-ploidy population would.
A nonzero probability of producing IBD diploid gametes ($\alpha_k > 0$) will
then further increase the inbreeding coefficient in the tetraploid and triploid
fraction of the population relative to their diploid progenitors
(\cref{s-fig:f1}).
Therefore, as long as diploids dominate, harboring some fraction of the gene
pool in polyploid individuals has a negligible effect on the rate of inbreeding
in the mixed-ploidy population as a whole, and we find that the evolution of
the inbreeding coefficient over time is well predicted by $1-e^{-t/2N_e}$,
where the inbreeding-effective population size is, to first order in $u$, given
by $(1-2u)N$ (\cref{s-sec:effsize}). This is just the expected number of
diploid individuals (to first order in $u$), highlighting that when diploids
dominate, polyploids have low fitness and do not contribute to the effective
population size.

@@

\subsection*{Establishment under recurrent migration}

\begin{figure}[t]
\centering
\includegraphics[width=0.8\linewidth]{/home/arzwa/dev/InfGenetics/doc/img/estwmigration-dr.pdf}
\caption{
    Establishment with recurrent migration.
    (A) Expected time until a population is established in the novel habitat
    for increasing rates of migration and different degrees of maladaptation
    ($\bar{z}_s$, the mean trait value in the source population). Results are
    shown for the case with $\alpha_k = 0$ for $k=2,3,4$. 
    (B) Proportion of simulation replicates in which tetraploids established.
    The dots connected by solid lines show simulation results with
    $\alpha_k=0$, whereas the dots connected by dashed lines
    show simulation results with $\alpha_2=1/2, \alpha_3=1/4$ and
    $\alpha_4=1/6$ (i.e. maximum $\alpha$).
    The solid horizontal lines mark $\frac{p_4\pi_4}{(p_4\pi_4 + p_2\pi_2)}$,
    where $p_2$ and $p_4$ are the establishment probabilities for a single
    diploid and a single tetraploid individual in the absence of migration (as
    in \cref{fig:est1}, but with unreduced gamete formation included). 
    The dashed horizontal line marks the proportion of tetraploid migrants
    (i.e. the proportion of tetraploids at equilibrium in the source
    population, $\approx 0.3$\%).
    The baseline predictions (horizontal lines) are based on $5\times 10^6$
    simulation replicates.
    All other results are based on 50.000 replicate simulations.
    We assume $\gamma=0.25$ and $u=v=0.05$.
    \label{fig:estwmig}}
\end{figure}

We next consider establishment in the new habitat when there is a continuous
influx of migrants ($m$ migrants per generation on average) coming from a large,
noninbred and predominantly diploid source population at cytotype equilibrium.
In this setting, establishment is certain to happen eventually, and we are
interested in the probability that a tetraploid population establishes before a
diploid one does.
More specifically, we are interested in how maladaptive gene flow and MCE
(which both increase with $m$) affect the relative establishment probability
and the time until establishment.

We hypothesized that two counteracting processes affect the probability of
autotetraploid establishment in this scenario.
On the one hand, increased migration will increase the probability that an
otherwise likely successful tetraploid migrant suffers from MCE in the early
generations while the population size is low, because migrants are likely to be
diploid.
On the other hand, tetraploids are more strongly reproductively isolated from
a typical migrant, so that a tetraploid subpopulation should be less prone to
maladaptive gene flow.
Hence, conditional on evading MCE, they should be able to adapt to the new
habitat at a rate which is not strongly affected by the migration rate.
This contrasts with diploids, which interbreed freely with maladapted
migrants, resulting in a pulling back of the trait mean towards that of the
source population.
However, MCE is expected to be stronger than the latter effect,
precluding the establishment of tetraploids as $m$ increases.
Lastly, as the mean trait value on the island increases in diploids
during adaptation, newly formed tetraploids on the island will, on average,
have more extreme phenotypes than diploid offspring when $\beta_4 > 1/2$, which
may also aid their establishment (irrespective of $m$).

\begin{figure}[t]
\centering
\includegraphics[width=\linewidth]{/home/arzwa/dev/InfGenetics/doc/img/estwmigration1d.pdf}
\caption{
    Evolution of the mean inbreeding coefficient and trait value across
    establishment replicates, averaged by population size, for increasing rates
    of migration ($m$) from the predominantly diploid source population. 
    All results are based on 1000 succesful establishment replicates.
    We assume equal equilibrium variance across ploidy levels and $\gamma=0.25,
    \theta=2, 2V=1$ and $u=v=0.05$.
    The $m=0$ simulations are seeded with a single diploid or tetraploid
    individual with trait value 0, and assume $u=v=0$.
\label{fig:estwmig2}}
\end{figure}

As expected, we find that the time to establishment (of a population of either
ploidy level) first decreases with increasing migration as a result of a larger
influx of potentially succesful migrants, but later increases with increasing
migration as a result of swamping by gene flow (\cref{fig:estwmig}A).
Importantly, the tetraploid establishment probability is considerably larger
than the expected proportion of tetraploid migrants over a large part of the
parameter range (\cref{fig:estwmig}B, black dashed line).
Furthermore, we find that for small migration rates, the probability that
tetraploids establish rather than diploids can be larger than expected based on
the establishment probabilities in the absence of migration (colored lines in
\cref{fig:estwmig}), depending on how maladapted migrant individuals are.
This suggests that maladaptive gene flow in diploids can increase the
probability of tetraploid establishment.
However, at the same time we find that the tetraploid establishment probability
declines monotonically with $m$, which highlights the negative effects of MCE.
Together, these results confirm the view that, while cytotype differences are an
effective barrier to maladaptive gene flow, this advantage to establishment is
outweighed by the negative effects of MCE.

Our simulations further show that the mechanism of unreduced gamete formation
(as determined by the $\alpha_2$ parameter) can affect the establishment
probability (\cref{fig:estwmig}B, dashed lines).
This is mainly because the phenotypic variance of a newly formed tetraploid is
increased by a factor $(1+\alpha_2)$, thereby increasing the chance that a
newly formed tetraploid is well-adapted to the marginal habitat.
The rate of double reduction ($\alpha_4$) has a more limited effect
(\cref{s-fig:suppdr}).

Established diploid populations are more inbred on average than established
tetraploids when migration is weak, but the difference is slight except when
there is no migration at all (\cref{fig:estwmig2}, top row).
For stronger migration ($m > 0.1$), the opposite holds.
This is a result of two interacting processes.
On the one hand, inbreeding is slower in tetraploids, so that during adaptation
and establishment from a single or limited number of barely inbred individuals, 
the inbreeding coefficient is expected to increase less rapidly conditional on
the event of tetraploid establishment.
On the other hand, migration mostly introduces unrelated diploids, which cross
more readily with diploids than tetraploids, so that migration reduces the
average relatedness in established diploid populations much more than in
established tetraploid populations.

Conditional on establishment, tetraploids have a higher trait mean both in the
absence of migration and for moderate to strong migration, but not when
migration is weak (\cref{fig:estwmig2}, bottom row).
In the absence of migration, this is a consequence of the reduced rate of
inbreeding and the resulting increased adaptive potential of tetraploids.
Another consequence of this is that, in the absence of migration, the average
trait value in tetraploids at low population size (conditional on eventual
establishment) is lower than that in diploids.
This is because, in order to adapt before genetic variation is exhausted,
tetraploids need not increase the trait mean as much as diploids do in the
first couple of generations.

It is clear from \cref{fig:estwmig2} that migration is not only maladaptive,
but also introduces genetic variation upon which selection can act, thereby
increasing the mean trait value at the time of establishment for both
tetraploids and diploids.
Interestingly, when migration is strong, tetraploids have markedly larger
phenotypic values than diploids ($m=1$ in \cref{fig:estwmig2}).
Diploids suffer strongly from maladaptive gene flow when the population size is
low, while tetraploids enjoy some reproductive isolation.
Furthermore, in these replicates, tetraploids tend to emerge and rise in
frequency at larger population sizes on the island, and hence tend to derive
from diploids that already experienced several generations of selection.
Neotetraploids deriving from parents with $z>0$ will have more extreme
phenotypes on average (see methods) and hence be better adapted.

\subsection*{Loss of self-incompatibility and assortative mating}

We now briefly examine the effects of prezygotic isolation mechanisms that may
be associated with polyploidization and that could alleviate MCE.
Plausible sources of prezygotic isolation include assortative mating by
cytotype and self-fertilization.

When polyploidization disrupts an existing SI system (see e.g.
\cite{robertson2011comparative,zenil2019,novikova2023}), we expect
that tetraploids suffer less from MCE, as an appreciable portion of their
ovules are now assured to be fertilized by diploid gametes, irrespective of the
composition of the population.
At the same time, we expect that accelerated inbreeding in selfing tetraploids
diminishes the adaptive advantage of tetraploids that we found above.
As shown in \cref{fig:selfing}, we find that when polyploidization is
associated with the loss of a SI system (i.e. when diploids are
self-incompatible, but tetraploids are not), tetraploids have a strongly
increases establishment probability.
This is the case even when the selfing rate $\sigma$ in tetraploids is zero (in
which case, under our modeling assumptions, there is only random selfing, i.e.
the \textit{realized} selfing rate in tetraploids is $1/N$).

Besides reducing MCE, selfing is also expected to simultaneously reduce the
effects of maladaptive gene flow, so we would expect that tetraploid
establishment is promoted with increasing migration rates. 
However, increased migration would at the same time alleviate the issue of
reproductive assurance in diploids, as a diploid migrant is more likely to find
a compatible mate.
We find that when the selfing rate is sufficiently high ($\ge 0.4$ in
\cref{fig:selfing}A), the relative establishment probability of tetraploids
increases with increasing migration rate.
In this regime, the increased strength of MCE due to increased migration and
increased reproductive assurance in diploids is compensated by the stronger
maladaptive gene flow experienced by diploids.

Self-incompatibility is clearly a strong disadvantage when colonizing a novel
habitat, as a self-incompatible population of size one can never reproduce.
However, even when diploids are self-compatible, polyploids may still have
increased rates of self-fertilization (for instance due to altered flower
morphology). As an illustration, \cref{fig:selfing}A also shows results where
diploids are assumed to be self-compatible with $\sigma_2=0$ (i.e. random
selfing, dashed transparent lines).
The tetraploid establishment probability is still markedly increased when
$\sigma_4 \ge 0.4$, and as for the simulations with self-incompatibility,
migration still promotes the probability of tetraploid establishment when the
selfing rate in polyploids is sufficiently large.

@@

The observation that polyploid populations tend to inhabit more extreme
habitats or occur at the edge of the range of their conspecific diploids
has spurred considerable interest among botanists and evolutionary biologists
\citep{rice2015,kolar2017,vandepeer2021,griswold2021,mortier2024}.
An important question is whether such patterns emerge because polyploids are
somehow more tolerant to extreme environmental conditions (i.e. they somehow
are intrinsically more fit than diploids at the range margin), or whether other
aspects of the population dynamics of mixed-ploidy populations may favor the
establishment of polyploid subpopulations in peripheral habitats.

In this study, we worked out the infinitesimal model for an additive polygenic
trait in autotetraploids and mixed-ploidy populations.
It would be very interesting to experimentally assess whether typical
quantitative traits in mixed-ploidy populations follow approximately the
mixed-ploidy infinitesimal model we outlined in this paper, and to estimate the
relevant model parameters (in particular the $\beta_3$ and $\beta_4$ values).

We used the mixed-ploidy infinitesimal model to study the establishment of
tetraploids in a marginal habitat by means of individual-based simulations. 
Assuming the trait to be under directional selection in the marginal habitat,
and migration of maladapted individuals from a predominantly diploid source,
we sought to determine the conditions under which tetraploids are more likely
to establish a stable population.
Importantly, we assume no intrinsic advantage of polyploids in the marginal
habitat, i.e. the expected fitness of a migrant individual is the same
regardless of the ploidy level.
Differences in establishment probabilities are hence caused solely by aspects
of autopolyploid genetics and the barrier to gene flow between subpopulations
of different ploidy levels.

@@

In the presence of maladaptive gene flow, a nascent tetraploid subpopulation
suffers from MCE, and although polyploids are more reproductively isolated from
a typical migrant (and hence suffer less maladaptive gene flow), MCE will
preclude polyploid establishment as the rate of migration grows.
Nevertheless, it is important to remark that, the probability of tetraploid
establishment in the marginal habitat can be an order of magnitude higher than
expected based on the frequency of tetraploid migrants (i.e. is roughly $O(u)$
instead of $O(u^2)$) when migration is sufficiently weak and maladaptation
sufficiently high.

@@

A major weakness of the present work is that we have ignored inbreeding
depression and dominance throughout.
Including dominance in the infinitesimal framework is already challenging for
diploids (requiring the tracking of 4-way identity coefficients;
\cite{barton2023}), and appears intractable for higher ploidy levels.
Nevertheless, autopolyploidy has important consequences whenever dominance is
relevant, as in the case of inbreeding depression.
Indeed, when inbreeding depression is due to recessive deleterious variation,
it should be less expressed in polyploids simply because, for the same allele
frequency, homozygous genotypes are much rarer in polyploids than in diploids
at Hardy-Weinberg equilibrium.
When there is indeed deleterious recessive variation, inbreeding during the
establishment process should therefore incur a higher fitness cost in diploids
relative to tetraploids, and hence further increase the probability of
tetraploid establishment.
In addition, inbreeding itself occurs at a slower rate in polyploids, further
decreasing inbreeding depression during establishment.

>>>>

    \textbf{Keywords:} polyploidy, adaptation, establishment, inbreeding,
    quantitative genetics 
\end{abstract}

@@

Hence, to explain the widespread occurrence of mixed-ploidy populations,
additional factors besides the continuous formation of polyploids through the
union of unreduced gametes need to be considered.
Firstly, chance establishment of tetraploids through drift could occur.
Indeed, the problem is somewhat analogous to the spread of underdominant
chromosomal rearrangements, where local establishment through genetic drift and
subsequent spreading in a subdivided population through local extinction and 
recolonization dynamics has been suggested as a plausible model
\citep{lande1985}.
However, MCE is quite strong in randomly mating mixed-ploidy populations, and
the population size has to be very small for local tetraploid establishment to
occur at an appreciable rate (\cite{rausch2005}, see also \cref{s-sec:mchain}).
Secondly, any form of \textit{prezygotic isolation} between cytotypes could
promote establishment of polyploid cytotypes by alleviating MCE.
Particularly relevant are assortative mating by cytotype  (for instance through
phenological differences across cytotypes, or differences in pollinators;
\cite{kolar2017}), self-fertilization \citep{rausch2005,novikova2023}, and
localized dispersal \citep{baack2005,kolar2017}.
Finally, selection may be invoked to explain the establishment of polyploids.
Tetraploids may have higher relative fitness than their diploid counterparts
due to reduced inbreeding depression \citep{ronfort1999,otto2000}, or due to being
better adapted to (changing) environmental conditions \citep{vandepeer2021}.

@@

In generation $t$, $M(t)$ migrant individuals arrive on an island (the new
habitat) joining $N^\ast(t)$ resident individuals, where $M(t)$ is Poisson
distributed with mean $m$. 
The migrant individuals are assumed to be unrelated to the resident
individuals.
After migration in generation $t$, the $N(t) = N^\ast(t) + M(t)$ individuals
reproduce sexually, and the offspring thus produced survives until the next
generation with a probability determined by their trait value.
In the basic model, random selfing is allowed (but see below for a model with
self-incompatibility).
We assume the trait is under directional selection, with fitness $w(z) =
e^{\gamma(z - \theta)}$, where $\gamma$ is the intensity of directional
selection and $\theta$ is the trait value for which the growth rate of the
island population becomes positive.
As the population becomes better adapted, eventually, some form of density
regulation must limit its growth.
We ignore density regulation throughout and focus on the initial establishment
phase, defining establishment as reaching a population of size 100.

@@

We evaluate the accuracy of the autotetraploid infinitesimal model as an
approximation to the evolution of a quantitative trait determined by $L$
additive loci.
We find that the infinitesimal model with inbreeding generally yields accurate
predictions for the evolution of the genetic variance when the number of loci
is sufficiently large ($L \ge 100$, say, \cref{fig:vztet,s-fig:fininf}).
Furthermore, we confirm that, in the absence of double reduction, the decay in
genetic variance due to inbreeding after a time $t$ is well-predicted by
$e^{-t/4N}$ (\cref{fig:vztet}A), as expected from the results of \cite{arnold2012}.
As predicted, double reduction (i.e. $\alpha_4 > 0$) leads to an immediate
increase in genetic variance (as it increases the segregation variance), but
leads to accelerated inbreeding, causing faster loss of genetic variance in the
long term (\cref{fig:vztet,s-fig:fininf}).
Simulations for the mixed-ploidy model further confirm the correctness of our
infinitesimal approximation (\cref{s-fig:vz}).

It is worth noting that, although inbreeding is slower in autotetraploids than
in diploids for the same population size, the tetraploid fraction of a
diploid-dominated mixed-ploidy population will have an equal or higher average
inbreeding coefficient (\cref{s-fig:f1}).
This is because in such a population, triploid and tetraploid individuals
mostly arise from gametes formed by diploid individuals, or by polyploid
individuals with very recent diploid ancestry (on average $1+u+2v$ generations ago
for tetraploids, and $1+\frac{2}{3}(u+2v)$ generations ago for triploids, see
\cref{s-sec:ttdip}).
A nonzero probability of producing IBD diploid gametes ($\alpha_k > 0$) will
then further increase the inbreeding coefficient in the tetraploid and triploid
fraction of the population relative to their diploid progenitors
(\cref{s-fig:f1}).
Therefore, as long as diploids dominate, harboring some fraction of the gene
pool in polyploid individuals has a negligible effect on the rate of inbreeding
in the mixed-ploidy population as a whole, and we find that the evolution of
the inbreeding coefficient over time is well predicted by $1-e^{-t/2N_e}$,
where the inbreeding-effective population size is, to first order in $u$, given
by $(1-2u)N$ (\cref{s-sec:effsize}). This is just the expected number of
diploid individuals (to first order in $u$), highlighting that when diploids
dominate, polyploids have low fitness and do not contribute to the effective
population size.

@@

\subsection*{Establishment with recurrent migration}

\begin{figure}[t]
\centering
\includegraphics[width=0.8\linewidth]{/home/arzwa/dev/InfGenetics/doc/img/estwmigration-dr.pdf}
\caption{
    Establishment with recurrent migration.
    (A) Expected time until a population is established in the marginal habitat
    for increasing rates of migration and different degrees of maladaptation
    ($\theta$). Results are shown for the case with $\alpha_k = 0$ for
    $k=2,3,4$. 
    (B) Proportion of simulation replicates in which tetraploids established.
    The dots connected by solid lines show simulation results with
    $\alpha_k=0$, whereas the dots connected by dashed lines
    show simulation results with $\alpha_2=1/2, \alpha_3=1/4$ and
    $\alpha_4=1/6$ (i.e. maximum $\alpha$).
    The solid horizontal lines mark $\frac{p_4\pi_4}{(p_4\pi_4 + p_2\pi_2)}$,
    where $p_2$ and $p_4$ are the establishment probabilities for a single
    diploid and a single tetraploid individual in the absence of migration (as
    in \cref{fig:est1}, but with unreduced gamete formation included). 
    The dashed horizontal line marks the proportion of tetraploid migrants
    (i.e. the proportion of tetraploids at equilibrium in the source
    population, $\approx 0.3$\%).
    The baseline predictions (horizontal lines) are based on $5\times 10^6$
    simulation replicates.
    All other results are based on 50.000 replicate simulations.
    We assume $\gamma=0.25$ and $u=v=0.05$.
    \label{fig:estwmig}}
\end{figure}

We next consider establishment in the new habitat when there is a continuous
influx of migrants ($m$ migrants per generation on average) coming from a large,
noninbred and predominantly diploid source population at cytotype equilibrium.
In this setting, establishment is certain to happen eventually, and we are
interested in the probability that a tetraploid population establishes before a
diploid one does.

We hypothesized that two counteracting processes affect the probability of
autotetraploid establishment in this scenario.
On the one hand, increased migration will increase the probability that an
otherwise likely successful tetraploid migrant suffers from MCE in the early
generations while the population size is low, because migrants are likely to be
diploid.
On the other hand, tetraploids are more strongly reproductively isolated from
a typical migrant, so that a tetraploid subpopulation should be less prone to
maladaptive gene flow.
Hence, conditional on evading MCE, they should be able to adapt to the new
habitat at a rate which is not strongly affected by the migration rate.
This contrasts with diploids, which interbreed freely with maladapted
migrants, resulting in a pulling back of the trait mean towards that of the
source population.
Lastly, as the mean trait value on the island increases in diploids
during adaptation, tetraploid offspring will have more extreme phenotypes on
average than diploid offspring when $\beta_4 > 1/2$, which may also aid their
establishment (irrespective of $m$).

\begin{figure}[t]
\centering
\includegraphics[width=\linewidth]{/home/arzwa/dev/InfGenetics/doc/img/estwmigration1d.pdf}
\caption{
    Evolution of the mean inbreeding coefficient and trait value across
    simulation replicates where diploids (blue) or tetraploids (orange)
    established eventually.
    Average $F$ and $z$ by population size are shown for increasing rates
    of migration ($m$) from the predominantly diploid source population. 
    All results are based on 1000 succesful establishment replicates.
    We assume equal equilibrium variance across ploidy levels and $\gamma=0.25,
    \theta=2, 2V=1$ and $u=v=0.05$.
    The $m=0$ simulations are seeded with a single diploid or tetraploid
    individual with trait value 0, and assume $u=v=0$.
\label{fig:estwmig2}}
\end{figure}

As expected, we find that the time to establishment (of a population of either
ploidy level) first decreases with increasing migration as a result of a larger
influx of potentially succesful migrants, but later increases with increasing
migration due to swamping by gene flow (\cref{fig:estwmig}A).
Importantly, the tetraploid establishment probability is considerably larger
than the expected proportion of tetraploid migrants over a large part of the
parameter range (\cref{fig:estwmig}B, black dashed line).
Furthermore, we find that for small migration rates, the probability that
tetraploids establish rather than diploids can be larger than expected based on
the establishment probabilities in the absence of migration (colored lines in
\cref{fig:estwmig}), depending on how maladapted migrant individuals are.
This suggests that maladaptive gene flow in diploids can increase the
probability of tetraploid establishment.
However, at the same time we find that the tetraploid establishment probability
declines monotonically with $m$, which highlights the negative effects of MCE.
Together, these results confirm the view that, while cytotype differences are an
effective barrier to maladaptive gene flow, this advantage to establishment is
outweighed by the negative effects of MCE.

Our simulations further show that the mechanism of unreduced gamete formation
(as determined by the $\alpha_2$ parameter) can affect the establishment
probability (\cref{fig:estwmig}B, dashed lines).
This is mainly because the phenotypic variance of a newly formed tetraploid is
increased by a factor $(1+\alpha_2)$, thereby increasing the chance that a
tetraploid migrant is well-adapted to the marginal habitat.
The rate of double reduction ($\alpha_4$) has a more limited effect
(\cref{s-fig:suppdr}).

Established diploid populations are more inbred on average than established
tetraploids when migration is weak, but the difference is slight except when
there is no migration at all (\cref{fig:estwmig2}, top row).
For stronger migration ($m > 0.1$), the opposite holds.
This is a result of two interacting processes.
On the one hand, inbreeding is slower in tetraploids, so that during adaptation
and establishment from a single or limited number of outbred individuals, 
the inbreeding coefficient is expected to increase less rapidly conditional on
the event of tetraploid establishment.
On the other hand, migration mostly introduces unrelated diploids, which cross
more readily with diploids than tetraploids, reducing the average relatedness
more strongly in established diploid than in tetraploid populations.

Conditional on establishment, tetraploids have a higher trait mean both in the
absence of migration and for moderate to strong migration, but not when
migration is weak (\cref{fig:estwmig2}, bottom row).
In the absence of migration, this is a consequence of the reduced rate of
inbreeding and the resulting increased adaptive potential of tetraploids.
Another consequence of this is that the average trait value in tetraploids at
low population size (conditional on eventual establishment) is lower than that
in diploids.
This is because, in order to adapt before genetic variation is exhausted,
tetraploids need not increase the trait mean as much as diploids do in the
first couple of generations.
The similar trait means for weak migration suggest that establishment depends
mostly on the chance pick of a well-adapted migrant (note that the $m=0$
simulations in \cref{fig:estwmig2} assume the initial migrant has $z=0$,
whereas the $m=0.01$ simulations assume the initial trait value to be
Gaussian).
When migration is strong, tetraploids have markedly larger trait values
than diploids ($m=1$ in \cref{fig:estwmig2}), showing that diploids suffer
strongly from maladaptive gene flow when the population size is low, while
tetraploids enjoy some reproductive isolation.
Furthermore, in these replicates, tetraploids tend to emerge and rise in
frequency at larger population sizes on the island, and hence tend to derive
from diploids that already experienced several generations of selection.
These neotetraploids, deriving from diploid parents with $z>0$, will have more
extreme phenotypes on average (see methods) and hence be better adapted.

\subsection*{Loss of self-incompatibility, selfing and assortative mating}

When polyploidization disrupts an existing SI system (see e.g.
\cite{robertson2011comparative,zenil2019,novikova2023}), we expect that
tetraploids suffer less from MCE, as some portion of their ovules are now
assured to be fertilized by diploid gametes, irrespective of the composition of
the population.
At the same time, we expect that accelerated inbreeding in selfing tetraploids
diminishes the adaptive advantage of tetraploids.
We find that when polyploidization is associated with the loss of a SI system
(i.e. when diploids are self-incompatible, but tetraploids are not),
tetraploids have a strongly increased establishment probability
(\cref{fig:selfing}).
This is the case even when the selfing rate $\sigma$ in tetraploids is zero (in
which case, under our modeling assumptions, there is only random selfing, i.e.
the \textit{realized} selfing rate in tetraploids is $1/N$).
Furthermore, we find that when the selfing rate is sufficiently high ($\ge 0.4$
in \cref{fig:selfing}A), the relative establishment probability of tetraploids
increases with increasing migration rate.
In this regime, the increased strength of MCE due to increased migration and
increased reproductive assurance in diploids is compensated by the stronger
maladaptive gene flow experienced by diploids.

Self-incompatibility is clearly a strong disadvantage when colonizing a novel
habitat, as a self-incompatible population of size one can never reproduce.
However, even when diploids are self-compatible, polyploids may still have
increased rates of self-fertilization (for instance due to altered flower
morphology).
For the sake of comparison, \cref{fig:selfing}A also shows results where
diploids are assumed to be self-compatible with $\sigma_2=0$ (i.e. random
selfing, dashed transparent lines).
The tetraploid establishment probability is still markedly increased when
$\sigma_4 \ge 0.4$, and as for the simulations with self-incompatibility,
migration still promotes the probability of tetraploid establishment when the
selfing rate in polyploids is sufficiently large compared to the diploid
selfing rate.

@@

The observation that polyploid populations tend to inhabit more extreme
habitats or occur at the edge of the range of their conspecific diploids
has spurred considerable interest among botanists and evolutionary biologists
\citep{rice2015,kolar2017,vandepeer2021,griswold2021,mortier2024}.
An important question is whether such patterns emerge because polyploids are
somehow more tolerant to extreme environmental conditions (i.e. they somehow
are intrinsically more fit than diploids in marginal habitats), or whether other
aspects of the population dynamics of mixed-ploidy populations may favor the
establishment of polyploid subpopulations.

In this study, we worked out the infinitesimal model for an additive polygenic
trait in autotetraploids and mixed-ploidy populations and used it to study the
establishment of tetraploids in a marginal habitat by means of individual-based
simulations.
Assuming the trait to be under directional selection in the marginal habitat,
and migration of maladapted individuals from a predominantly diploid source,
we sought to determine the conditions under which tetraploids are more likely
to establish a stable population.

Importantly, we assume no intrinsic advantage or disadvantage of polyploids in
the marginal habitat, i.e. the expected fitness of a migrant individual is the
same regardless of the ploidy level.
Differences in establishment probabilities are hence caused solely by aspects
of autopolyploid genetics and the barrier to gene flow between subpopulations
of different ploidy levels.
This is undoubtedly unrealistic.
For instance, neopolyploids are likely to suffer intrinsic fertility issues due
to meiotic irregularities associated with multivalent formation
\citep{bomblies2016,novikova2023}.
Furthermore, in our simulations, we have focused on the case where the
equilibrium genetic variance is constant across cytotypes ($\beta_4=\sqrt{1/2}$
in our model).
It would be very interesting to experimentally assess whether typical
quantitative traits in mixed-ploidy populations follow approximately the
mixed-ploidy infinitesimal model we outlined in this paper, and to actually
estimate how the variance scales across ploidy levels.
In addition, we have assumed a relatively high and constant rate of unreduced
gamete formation $u$ and triploid fertility $v$ in all our simulations (5\%),
whereas these are known to be variable across the population, and at least in
part genetically determined \citep{kreiner2017,clo2022c}.
We ignore such complications, and hence do not take the actual establishment
probabilities very serious, focusing instead on how migration load and
prezygotic isolation affect the tetraploid establishment probability.

@@

In the presence of maladaptive gene flow, a nascent tetraploid subpopulation
suffers from MCE, and although polyploids are more reproductively isolated from
a typical migrant (and hence suffer less maladaptive gene flow), MCE will
increasingly hamper the establishment of tetraploids as the rate of migration
grows.
Nevertheless, it is important to remark that despite MCE, the probability of
tetraploid establishment in the marginal habitat can be an order of magnitude
higher than expected based on the frequency of tetraploid migrants (i.e. is
roughly $O(u)$ instead of $O(u^2)$) when migration is sufficiently weak and
maladaptation sufficiently high.

@@

A major weakness of the present work, and an important caveat, is that we have
ignored inbreeding depression and dominance throughout.
Including dominance in the infinitesimal framework is already challenging for
diploids (requiring the tracking of four-way identity coefficients;
\cite{barton2023}), and appears intractable for higher ploidy levels.
However, autopolyploidy has important consequences whenever dominance is
relevant, as in the case of inbreeding depression \citep{ronfort1999}.
Indeed, when inbreeding depression is due to recessive deleterious variation,
it is expected to be less expressed in neotetraploids because homozygous genotypes
should be much rarer than in their diploid parents.
When there is indeed deleterious recessive variation, inbreeding during the
establishment process should therefore incur a higher fitness cost in diploids
relative to tetraploids, and hence further increase the probability of
tetraploid establishment.
How this plays out depends on the \textit{rate} of inbreeding, which will
differ between cytotypes. 
In outcrossing populations, inbreeding occurs at a slower rate in tetraploids,
further decreasing inbreeding depression and aiding tetraploid establishment.
However, when polyploidization is associated with increased selfing (as when it
disrupts an existing SI system), increased inbreeding depression in
autotetraploids may prevent their establishment.

Dominance and inbreeding depression may strongly affect the complicated
relationship between selfing, migration load and tetraploid establishment
probability.
\cite{griswold2021} studied the case where local fitness is determined by a
single biallelic locus, and investigated the interaction between inbreeding
depression and migration load (where inbreeding depression is modeled as a
fixed fitness reduction in offspring produced by selfing).
In his model, inbreeding depression is different between cytotypes (assuming
stronger inbreeding depression in diploids), so that tetraploids are able to
produce more offspring through selfing relative to diploids, who have to rely
more on outcrossing.
However, outcrossing incurs maladaptive gene flow, and thereby puts the
diploids at a disadvantage. 
He found that autotetraploids can establish when adaptation in the
peripheral habitat is conferred by recessive alleles (so that migration load is
expressed when migrant alleles are rare) and when inbreeding depression in
tetraploids is lower than in diploids.
It would be very interesting to combine the infinitesimal framework with some
form of inbreeding depression to investigate in a more realistic model whether
the combination of maladaptive migration and differential inbreeding depression
could explain the prevalence of polyploid subpopulations at range edges.
